On the Synchronization Power of the OpenFlow Data Plane

### 2015-05-29 inband-sync.tex
first 6-page version

While the perspective of \emph{logically centralized} control plane
offered by SDN is intuitive and attractive,
there is a wide consensus that
the control plane should be  \emph{physically distributed}.
First, in order to provide high availability,
%as well as a minimal degree of
%fault-tolerance,
%[[PK fault-tolerance is necessary to provide hight availability?]]
controllers should be redundant~\cite{onix,stn,onos,elasticon}: a failure 
of one controller can be masked by other controllers.
Second, it has also been proposed to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication-intensive
%control-plane
data-plane
events close to their origin~\cite{sharon,devoflow,kandoo,jukka,disco}.
Third, larger SDNs are likely to be operated by multiple administrators~\cite{stn} or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).
\emph{Synchronizing}
the distributed controllers and manipulating the network state \emph{consistently}
are non-trivial tasks.~\cite{cpc,sharon,cap-theorem}

Consider, for example, the problem of
consistent installation of new forwarding policies, stipulating routes
that packets of different header spaces should follow across the
network~\cite{network-update,podc15,roger-hotnets,correct,stn}.
Installing \emph{conflicting} forwarding rules, e.g., rules of the same priority defined over non-disjoint
flow spaces may lead to pathological network behavior (loops,
blackholes, routes bypassing a firewall, etc.).~\cite{hotnets14update,roger-hotnets}
Similarly, installing diverging load-balancing policies may,
when combined, \emph{increase} the load~\cite{log-cent}.
To render things more difficult, controllers may also fail,
even before their updates have been completed.

The concurrent computing literature offers
a wide range of synchronization abstractions~\cite{HS08-book}.
E.g., the popular \emph{transactional
  memory} abstraction~\cite{stm-st95,tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in atomic \emph{transactions} with
all-or-nothing semantics.

@@

\paragraph{Contributions}
This paper applies the principle of atomicity in concurrent computing
to distributed SDN control planes.
%
In particular, we propose synchronization constructs which
allow a controller to represent multiple configuration commands on
the data plane as an atomic transaction.
%
If none of the transaction commands \emph{conflicts}  with the current
configuration, where conflicts can be
defined in a general, application-specific manner, the transaction appears to be executed atomically.
Otherwise, the transaction is aborted in its entirety and does not
affect neither traffic nor other controllers.

We promote to implement our synchronization constructs \emph{inband},
on the OpenFlow switch.
An inband implementation allows us to efficiently alleviate the problems related to
coordinating controllers via a control-plane
out-of-band network, in the presence of asynchrony and failures.
Indeed, the inherent costs  of such protocols are often considered too high, both in
terms of the necessary computability assumptions about the underlying
system~\cite{FLP85,CHT96,cap-theorem}, and the amount of communication
complexity~\cite{Lam98,Lam06}.
In contrast, our \emph{in-band} solution allows the
controllers to solve fundamental agreement tasks in just one message
exchange with the data plane, tolerating asynchrony and failures of
any number of controllers.

Interestingly, our mechanisms can be implemented using the \emph{standard OpenFlow} 
protocol, and do not require any extensions as assumed and postulated in recent literature (e.g.,
STN~\cite{stn} or NetPaxos~\cite{netpaxos}). 
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{compute,reclaim},
as well as to possible high-level concurrency objects and 
language abstractions.~\cite{pyretic}

%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}

Concretely, in our implementation, a part of the data plane configuration space is
used as a shared memory storing information about  contention and
conflicts between distributed controllers.
A transaction can then contain standard OpenFlow control operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
Our synchronization primitives allow the controllers to define complex notions
of conflicts that can be used for coordinating configuration updates operating 
over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is
important, e.g., for load-balancing applications.

@@

We then show that, using our abstraction, the controllers can
solve several important synchronization problems.
We describe an implementation of the fundamental
\emph{compare-and-set} (CAS) primitive,
which in turn can be used to solve consensus and
implement a generic replicated control-plane service~\cite{Her91,Lam98} in a consistent and
fault-tolerant way, which became a mainstream building block in
modern system design~\cite{bigtable,spanner,zookeeper}.
We then discuss a simple CAS-based concurrent policy-update mechanism.
Finally, we use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{cpc}. \petr{We do not seem to discuss the latter?}

@@

\paragraph{Control and data planes}
Software-defined networking decouples the \emph{control plane}, software
controllers that configure the network, from the \emph{data plane}, the
devices that implement the network.
%
In OpenFlow~\cite{of-spec}, the \emph{de facto} standard SDN protocol
today, the data plane consists of \emph{OpenFlow switches}.
In a nutshell, the configuration of an OpenFlow switch is a set of
\emph{rules}.
A rule is essentially a \emph{match-action} pair:
the match part of a rule identifies the space of packet headers that are
subject to the rule (e.g., all packets to a specific destination) and
the action part identifies how the switch should process the matched
packets (e.g., forward them to a specific port).

@@

\paragraph{Configuration space}
%
We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules),
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.
Flow entries in the meta-configuration can be referred to as shared
memory locations, provided a total addressing scheme.

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data-plane
traffic. One way to achieve this is to set a low priority to
the meta flow entries and ensure that the regular flow entries
maintain higher priority levels.
Alternatively, we could leverage the standard ability of maintaining multiple
flow tables at an OpenFlow switch, and use a separate flow table for
the meta-configuration,  making sure that the
\emph{pipelining} mechanism~\cite{of-spec} will never involve this
flow table.

@@

\paragraph{The id-claimer}

@@

In order to check if an identifier is claimed, a controller tries to add 
a new entry setting the flag \texttt{\checko}, and using a different
action ($2$ instead of $1$),
thereby inflicting a failure in case an entry with overlapping match value exists.
If the check succeeds, we delete the entry.
Con sidering, $m_1$ and $m_2$ from the example above,
with respective actions $a_1=1$ and $a_2=2$,  they overlap
with different actions iff $x_1=x_2$. Therefore, if $m_1$ is used in \claimcheck after $m_2$ used in the claim operation,
a conflict is detected iff $x_1=x_2$, as desired.

Note that for space reasons, when checking for claims, also the deletion of the added tester flow entry is included,
which however has no effect on claims and checks of other controllers; hence, the deletion command does not
have to be sent as part of a bundle. Therefore, the \texttt{\claimcheck} operation is atomic by nature,
similar to the \texttt{claim} and \texttt{unclaim} operations, and is implemented as essentially one command.

@@

\paragraph{Write and compare}
The memory \memwrite and \compare
 operations can be implemented using similar techniques.
 As can be seen in Algorithms \ref{alg:write}-\ref{alg:compare},
  the match part of the added flow entry is used as the memory address and the action as
  the written value. Values written to the same address will replace one another,
following the command definition to replace the old flow entry with a new one in case they
share the same match parts.

In order to check if a value is written in a given address,
we try to add it the same way as in the \memwrite operation,
but we also set the \texttt{\checko} flag. If the value is not there,
our action differs from the action of the existing entry,
thereby inflicting a failure.
In case the value is there, our flow entry replaces the existing flow entry
with an identical one.

@@

We now discuss several applications of our transactional interface.

\textbf{Stefan: Give intro
}%very powerful primitive => any sequential program in repliated way can be imlemented, see also STN paper where we assume we can do this.

\paragraph{Compare-and-Set}
The pathological scenario described in the example in
Figure~\ref{fig:CAS-example} can be resolved easily if we put
the policy update in our transactional $\execatomic$ construct together
 with a \emph{compare-and-set} (CAS) operation on the policy identifier.
A $\cas(\addr,\textit{old},\textit{new})$ operation on a memory location $\addr$
checks if the content of $\addr$ is $\textit{old}$ and if so,
replaces it with $\textit{new}$ and returns $\true$ (the CAS
\emph{succeeds}), otherwise it returns $\true$ (the CAS
\emph{fails} which causes an abort of
the invoking transaction).
It is straightforward to implement the CAS operation by executing
$\textit{\compare}(\addr, old)$ followed by $\textit{\memwrite}(\addr,
new)$ within the same bundle.

@@

CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data-plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and, more generally,
to implement any generic replicated state machine~\cite{Her91,Lam98}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
%which will be instrumental in the
%next section.

@@

\paragraph{Policy Updates}

As a second example, let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.

Let us assume that the composition predicate is encoded in a user-specific
$\textsc{\ufunc}()$ function~\cite{cpc}.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
% \liron{the term composition doesn't make sense now with the new user update function... }

\liron{we can explain how a normal controller code can be instrumented to use our paradigm}
\petr{did not get the comment}

@@

\petr{what does it mean for an update function to be ``unsafe''?}

@@

Note that in Algorithm~\ref{alg:simple-update}, policies identifiers
grow without bound.
However, as was observed in~\cite{cpc}, the number of currently
used identifiers can be made proportional to the number of concurrent
controllers.
%
\hide{
It takes a set of rules $U$ (not yet a policy)and proceeds as follows: first, we seek to
 obtain a unique \emph{id}. FIXME: to be continued...
 \textbf{LS: I am not sure if I need to tell every step of the story or maybe it best to explain the main dif from the previous one similarly to what I just wrote next}.
}
%

@@

@@

%\paragraph{One-Shot Update} FIXME: desribe how to implement one-shot update
%
%\paragraph{STN}
%
%\textbf{TODO}
%

@@

\noindent\textbf{Distributed Computing.}
There is a long tradition of defining correctness of a concurrent system via
an equivalence to a sequential one~\cite{Pap79-serial,Lam79,HW90}.  The notion
of sequentially composable histories is reminiscent of
linearizability~\cite{HW90}, where a history of operations concurrently
applied by a collection of processes is equivalent to a history in which the
operations are in a sequential order, respecting their real-time precedence.
In contrast, our sequentially composable histories impose requirements not
only on high-level invocations and responses, but also on the way the traffic
is processed. We require that the committed policies constitute a
conflict-free sequential history, but, additionally,  we expect that each
\emph{path} witnesses only a prefix of this history, consisting of all
requests that were committed before the path was initiated.
%
The transactional interface exported by the CPC abstraction is inspired by the
work on speculative concurrency control using software transactional memory
(STM)~\cite{stm-st95}.
%, thus the  term \emph{software transactional networking}.
Our interface is however intended to model realistic network
management operations, which makes it simpler than recent
dynamic STM models~\cite{dstm}.
%In particular, the sets of rules to be installed by a policy update do not depend on the state of the
%network.

Extending the interface to dynamic policies that adapt their behavior based on
the current network state sounds like a promising research direction.  On the
other hand, our criterion of sequential composability is more complex than
traditional STM correctness properties in that it imposes restrictions not only
on the high-level interface exported to the control plane, but also on the
paths taken by the data-plane packets.
Also, we assumed that controllers are subject to failures, which is usually not
assumed by STM implementations.
}

@@

\section{Concluding Remarks}\label{sec:conclusion}

While much progress has been made over the
last years in the development of more intuitive programming
languages for SDNs (e.g., Frenetic, Pyretic, Merlin), the
question of which synchronization abstractions to provide
has received much less attention.
%
In this paper, we describe transactional abstractions that
can be implemented using standard OpenFlow features.
%
Our work complements existing research on the design of
distributed control planes, and also provides some missing links,
e.g., for the transactional approach taken by STN~\cite{stn}.
We also hope that our work can nourish the discussion of
which synchronization objects we can and want to implement
in an SDN.

Even though we believe that we propose efficient and consistent
transactional framework for in-band synchronization, we do not claim
that we offer an exhaustive set of synchronization primitives. Indeed,
one may explore standard OpenFlow features to enable more powerful
primitives. For instance, \emph{cookies}~\cite{openflow} can be used for modifying
flow tables, based on more sophisticated conditions than the default
``exact match''.
Also, the standard pipelining feature allows the switches to maintain multiple flow
tables which can be traversed in arbitrary patterns, defining the sets
of corresponding actions.
Exploring the power of such features is left to future work.

{
\bibliographystyle{abbrv}
\bibliography{references}  % main.bib is the name of the Bibliography in this case
}

@@

We first show that they can be used to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
recent proposal for a consistent composition system for forwarding policies~\cite{cpc}.
We then use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.
We then explore how more general classes of policies can be composed
\emph{on-the-switch}.

>>>>

While the perspective of \emph{logically centralized} control plane
offered by SDN is intuitive and attractive,
there is a wide consensus that
the control plane should be  \emph{physically distributed}.
First, in order to provide high availability,
%as well as a minimal degree of
%fault-tolerance,
%[[PK fault-tolerance is necessary to provide hight availability?]]
controllers should be redundant~\cite{onix,onos,elasticon}: a failure 
of one controller can be masked by other controllers.
Second, it has also been proposed to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication-intensive
%control plane
data plane
events close to their origin~\cite{devoflow,kandoo}.
Third, larger SDNs are likely to be operated by multiple administrators~\cite{stn} or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory}.

@@

Today, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed systems
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).
\emph{Synchronizing}
the distributed controllers and manipulating the network state \emph{consistently}
are non-trivial tasks.~\cite{sharon}

Consider, for example, the problem of
consistent installation of new forwarding policies, stipulating routes
that packets of different header spaces should follow across the
network~\cite{network-update,roger-hotnets,podc15}.
Installing \emph{conflicting} forwarding rules, e.g., rules of the same priority defined over non-disjoint
flow spaces may lead to pathological network behavior (loops,
blackholes, routes bypassing a firewall, etc.).~\cite{hotnets14update,roger-hotnets}
Similarly, installing diverging load-balancing policies may,
when combined, \emph{increase} the load.
To render things more difficult, controllers may also fail,
even before their updates have been completed.

The concurrent computing literature offers
a wide range of synchronization abstractions.
E.g., the popular \emph{transactional
  memory} abstraction~\cite{tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in atomic \emph{transactions} with
all-or-nothing semantics.

@@

\noindent\textbf{Contributions}
This paper applies the principle of atomicity in concurrent computing
to distributed SDN control planes.
%
In particular, we propose synchronization constructs which
allow a controller to represent multiple configuration commands on
the data plane as an atomic transaction.
%
If none of the transaction commands \emph{conflicts}  with the current
configuration, where conflicts can be
defined in a general, application-specific manner, the transaction appears to be executed atomically.
Otherwise, the transaction is aborted in its entirety and does not
affect neither traffic nor other controllers.

We promote to implement our synchronization constructs \emph{inband},
on the OpenFlow switch.
An inband implementation allows us to efficiently alleviate the problems related to
coordinating controllers via a control plane
out-of-band network, in the presence of asynchrony and failures.
Indeed, the inherent costs  of such protocols are often considered too high, both in
terms of the necessary computability assumptions about the underlying
system~\cite{FLP85}, and the amount of communication
complexity~\cite{Lam06}.
In contrast, our \emph{in-band} solution allows the
controllers to solve fundamental agreement tasks in just one message
exchange with the data plane, tolerating asynchrony and failures of
any number of controllers.

Interestingly, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol. 
%
%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}
%
Concretely, in our implementation, 
as part of the data plane configuration space, 
we propose to use a shared memory storing information about  contention and
conflicts between distributed controllers.
A transaction can then contain standard OpenFlow control operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
Our synchronization primitives allow the controllers to define complex notions
of conflicts that can be used for coordinating configuration updates operating 
over \emph{independent flow spaces}, but where dependencies are introduced
in the mapping of the flows to the physical infrastructure. This is
important, e.g., for load-balancing applications.

@@

We then show that, using our abstraction, the controllers can
solve several important synchronization problems.
We describe an implementation of the fundamental
\emph{compare-and-set} (CAS) primitive,
which in turn can be used to solve consensus and
implement a generic replicated control plane service~\cite{Her91} in a consistent and
fault-tolerant way: a mainstream building block in
modern system design.
We also discuss a simple CAS-based concurrent policy-update mechanism,
and also use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{stn}. \petr{We do not seem to discuss the latter?}

@@

\noindent\textbf{Control and data planes}
Software-defined networking decouples the \emph{control plane}, software
controllers that configure the network, from the \emph{data plane}, the
devices that implement the network.
%
In OpenFlow, the \emph{de facto} standard SDN protocol
today, the data plane consists of \emph{OpenFlow switches}.
In a nutshell, the configuration of an OpenFlow switch is a set of
\emph{rules}.
A rule is essentially a \emph{match-action} pair:
the match part of a rule identifies the space of packet headers that are
subject to the rule (e.g., all packets to a specific destination) and
the action part identifies how the switch should process the matched
packets (e.g., forward them to a specific port).

@@

\noindent\textbf{Configuration space}
%
We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data plane traffic (e.g., forwarding rules),
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.
Flow entries in the meta-configuration can be referred to as shared
memory locations, provided a total addressing scheme.

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data plane
traffic. One way to achieve this is to set a low priority to
the meta flow entries and ensure that the regular flow entries
maintain higher priority levels.
Alternatively, we could leverage the standard ability of maintaining multiple
flow tables at an OpenFlow switch, and use a separate flow table for
the meta-configuration,  making sure that the
\emph{pipelining} mechanism will never involve this
flow table.

@@

\noindent\textbf{The id-claimer}

@@

In order to check if an identifier is claimed, a controller tries to add 
a new entry setting the flag \texttt{\checko}, and using a different
action ($2$ instead of $1$),
thereby inflicting a failure in case an entry with overlapping match value exists.
If the check succeeds, we delete the entry.
Considering $m_1$ and $m_2$ from the example above,
with respective actions $a_1=1$ and $a_2=2$,  they overlap
with different actions iff $x_1=x_2$. Therefore, if $m_1$ is used in the \claimcheck after $m_2$ was used in the claim operation,
a conflict is detected iff $x_1=x_2$, as desired.

Note that in order to save space, when checking for claims, also the deletion of the added tester flow entry is included,
which however has no effect on claims and checks of other controllers; hence, the deletion command does not
have to be sent as part of a bundle. Therefore, the \texttt{\claimcheck} operation is atomic by nature,
similar to the \texttt{claim} and \texttt{unclaim} operations, and is implemented as essentially one command.

@@

\noindent\textbf{Write and compare}
The memory \memwrite and \compare
 operations can be implemented using similar techniques.
 As can be seen in Algorithms \ref{alg:write}-\ref{alg:compare},
  the match part of the added flow entry is used as the memory address and the action as
  the written value. Values written to the same address will replace one another,
following the command definition to replace the old flow entry with a new one in case they
share the same match parts.

In order to check if a value is written at a given address,
we try to add it the same way as in the \memwrite operation,
but we also set the \texttt{\checko} flag. If the value is not there,
our action differs from the action of the existing entry,
thereby inflicting a failure.
In case the value is there, our flow entry replaces the existing flow entry
with an identical one.

@@

We believe that our synchronization mechanisms 
can be useful in many situations. 
For example, the pathological scenario described in 
Figure~\ref{fig:CAS-example} can be resolved easily if we put
the policy update in our transactional $\execatomic$ construct together
 with a \emph{Compare-and-Set} (CAS) operation on the policy identifier.
A $\cas(\addr,\textit{old},\textit{new})$ operation on a memory location $\addr$
checks if the content of $\addr$ is $\textit{old}$ and if so,
replaces it with $\textit{new}$ and returns $\true$ (the CAS
\emph{succeeds}), otherwise it returns $\false$ (the CAS
\emph{fails} which causes an abort of
the invoking transaction).
It is straightforward to implement the CAS operation by executing
$\textit{\compare}(\addr, old)$ followed by $\textit{\memwrite}(\addr,
new)$ within the same bundle.

@@

More generally, CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and allows us
to implement any generic replicated state machine~\cite{Her91}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
As such, CAS can also be used to implement STN~\cite{stn},
by providing the missing synchronization object.

%which will be instrumental in the
%next section.

@@

Our mechanism can also be used as a generic template to
implement concurrent policy compositions of a user-specific
$\textsc{\ufunc}()$ function~\cite{stn}.
That is, let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
% \liron{the term composition doesn't make sense now with the new user update function... }

%\liron{we can explain how a normal controller code can be instrumented to use our paradigm}
%\petr{did not get the comment}

@@

%\petr{what does it mean for an update function to be ``unsafe''?}

@@

%Note that in Algorithm~\ref{alg:simple-update}, policies identifiers
%grow without bound.
%However, as was observed in~\cite{stn}, the number of currently
%used identifiers can be made proportional to the number of concurrent
%controllers.
%
\hide{
It takes a set of rules $U$ (not yet a policy)and proceeds as follows: first, we seek to
 obtain a unique \emph{id}. FIXME: to be continued...
 \textbf{LS: I am not sure if I need to tell every step of the story or maybe it best to explain the main dif from the previous one similarly to what I just wrote next}.
}
%

@@

In other words, Algorithm~\ref{alg:simple-update} can also be used
to render our naive load-balancing algorithm (Algorithm~\ref{alg:naive-lb}) from Section~\ref{sec:motivation}
consistent, without any modifications.

@@

%\noindent\textbf{One-Shot Update} FIXME: desribe how to implement one-shot update
%
%\noindent\textbf{STN}
%
%\textbf{TODO}
%

@@

\noindent\textbf{Distributed Computing.}
There is a long tradition of defining correctness of a concurrent system via
an equivalence to a sequential one~\cite{Pap79-serial,Lam79,HW90}.  The notion
of sequentially composable histories is reminiscent of
linearizability~\cite{HW90}, where a history of operations concurrently
applied by a collection of processes is equivalent to a history in which the
operations are in a sequential order, respecting their real-time precedence.
In contrast, our sequentially composable histories impose requirements not
only on high-level invocations and responses, but also on the way the traffic
is processed. We require that the committed policies constitute a
conflict-free sequential history, but, additionally,  we expect that each
\emph{path} witnesses only a prefix of this history, consisting of all
requests that were committed before the path was initiated.
%
The transactional interface exported by the STN abstraction is inspired by the
work on speculative concurrency control using software transactional memory
(STM)~\cite{stm-st95}.
%, thus the  term \emph{software transactional networking}.
Our interface is however intended to model realistic network
management operations, which makes it simpler than recent
dynamic STM models~\cite{dstm}.
%In particular, the sets of rules to be installed by a policy update do not depend on the state of the
%network.

Extending the interface to dynamic policies that adapt their behavior based on
the current network state sounds like a promising research direction.  On the
other hand, our criterion of sequential composability is more complex than
traditional STM correctness properties in that it imposes restrictions not only
on the high-level interface exported to the control plane, but also on the
paths taken by the data plane packets.
Also, we assumed that controllers are subject to failures, which is usually not
assumed by STM implementations.
}

@@

\section{Conclusion}\label{sec:conclusion}

This paper initiated the study of mechanisms which allow distributed controllers 
to synchronize their possibly concurrent actions. We believe that our in-band 
approach is natural and attractive as it can avoid the 
complexities and limitations of distributed out-of-band agreement protocols.
Moreover, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions.~\cite{pyretic}

%,
%nor on
%nor do we make any assumptions on packet orderings.
%and do not rely on any
%
%.

\begin{comment}
While much progress has been made over the
last years in the development of more intuitive programming
languages for SDNs (e.g., Frenetic, Pyretic, Merlin), the
question of which synchronization abstractions to provide
has received much less attention.
%
In this paper, we describe transactional abstractions that
can be implemented using standard OpenFlow features.
%
Our work complements existing research on the design of
distributed control planes, and also provides some missing links,
e.g., for the transactional approach taken by STN~\cite{stn}.
We also hope that our work can nourish the discussion of
which synchronization objects we can and want to implement
in an SDN.

Even though we believe that we propose efficient and consistent
transactional framework for in-band synchronization, we do not claim
that we offer an exhaustive set of synchronization primitives. Indeed,
one may explore standard OpenFlow features to enable more powerful
primitives. For instance, \emph{cookies}~\cite{openflow} can be used for modifying
flow tables, based on more sophisticated conditions than the default
``exact match''.
Also, the standard pipelining feature allows the switches to maintain multiple flow
tables which can be traversed in arbitrary patterns, defining the sets
of corresponding actions.
Exploring the power of such features is left to future work.
\end{comment}

{
%\begin{footnotesize}
\bibliographystyle{abbrv}
\bibliography{references}  % main.bib is the name of the Bibliography in this case
%\end{footnotesize}
}

@@

We first show that they can be used to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
recent proposal for a consistent composition system for forwarding policies~\cite{stn}.
We then use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.
We then explore how more general classes of policies can be composed
\emph{on-the-switch}.
